---
title: "Wings to Fly"
subtitle: "Analyzing Delta Airlines: A Study of Late Arrivals Across U.S. Airports"
author: "Felix Jena"
format: 
  html:
    self-contained: true
    page-layout: full
    title-block-banner: true
    toc: true
    toc-depth: 3
    toc-location: body
    number-sections: false
    html-math-method: katex
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    code-copy: hover
    code-tools:
        source: false
        toggle: true
        caption: "See code"
execute: 
  warning: false
editor_options: 
  chunk_output_type: inline
---
### Research Questions:
Which Origin Airport Minimizes the Chances of Late Arrival?\
Which Destination Airport is the Worst for Arrival Delays?

### Libraries being used:
```{r}
library(tidyverse)
library(nycflights13)
```

### Filtering for Delta Airlines
```{r}

delta_flights <- flights %>%
  filter(carrier == "DL", !is.na(arr_delay))

```

### Which Origin Airport Minimizes the Chances of Late Arrival?
```{r}
# If Arrival is late
delta_flights <- delta_flights %>%
  mutate(late_arrival = if_else(arr_delay > 0, "Late", "On Time"))

# Proportion late arrivals summary
origin_late_prop <- delta_flights %>%
  group_by(origin) %>%
  summarize(total_flights = n(),
            late_arrivals = sum(late_arrival == "Late"),
            proportion_late = late_arrivals / total_flights) %>%
  arrange(proportion_late)


ggplot(origin_late_prop, aes(x = reorder(origin, -proportion_late), y = proportion_late * 100, fill = origin)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Percentage
  labs(title = "Late Arrivals by Origin (Delta Airlines)",
       x = "Airport",
       y = "Late Arrivals") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels


```
The best airport for minimizing the chances of a late arrival is JFK (John F. Kennedy), where only 30% of the flights arrive late.\
LGA (LaGuardia) comes second, with a 35% late arrival rate.\
EWR (Newark) has the highest proportion of late arrivals at 40%, meaning it’s less reliable in avoiding delays.\


### Which Destination Airport is the Worst for Arrival Delays?
```{r}
# Summarize Average arrival delay
dest_avg_delay <- delta_flights %>%
  group_by(dest) %>%
  summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(avg_arr_delay))


ggplot(dest_avg_delay, aes(x = reorder(dest, -avg_arr_delay), y = avg_arr_delay)) +
  geom_bar(stat = "identity", fill = "steelblue", color = "black") +
  labs(title = "Average Arrival Delay by Destination (Delta Airlines)",
       x = "Destination Airport",
       y = "Average Arrival Delay (minutes)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotates x-axis labels

```
Jackson Hole Airport (JAC) has the worst average arrival delay at 11 minutes, making it the destination with the most consistent delays among the analyzed airports.\
San Antonio International Airport (SAT) follows closely with an average delay of 8.08 minutes, and Hartsfield-Jackson Atlanta International Airport (ATL) ranks third with an average delay of 7.42 minutes.\
On the opposite side of the spectrum, Nashville International Airport (BNA) and Indianapolis International Airport (IND) have the best on-time performance, with average early arrivals of -27 minutes and -26.5 minutes, respectively.\

### Conclusion:
If you are flying Delta Airlines and want to minimize your chances of a late arrival, choose JFK as your airport.\
Jackson Hole (JAC) and San Antonio (SAT) have the worst average arrival delays, with 11 minutes and 8.08 minutes. Hartsfield-Jackson Atlanta (ATL) also ranks among the worst with an average delay of 7.42 minutes. Avoid these destinations if minimizing delays is important for your travel plans.



